---
title: "Feature insights into Somalia’s severe acute malnutrition admissions"
subtitle: "A time series analysis spanning from January 2019 to November 2024"
author: "Tomás Zaba"
date: today
format: revealjs
---

## Context {.smaller} 

::: {.incremental}
+ Seasonality/seasonal refers to recurring patterns that occurs at specific times of the year. 
+ It has impact on the underlying and immediate factors that drives acute malnutrition (AMN). 
+ It is known that AMN has a seasonal variation, however limited evidence is available to substantiate that.  
  - Stakeholders often rely on assumptions of an AMN seasonal peak during the food security lean season.  
    - Nutrition surveys are routinely timed to coincide with the lean season each year. 
::: 

## Somalia's Food Security Seasonal Calendar {.smaller} 

::: {.incremental}
```{mermaid}
%%| label: seasonal-calendar
%%| fig-cap-location: bottom
%%| fig-height: 4
%%| fig-width: 12
gantt
    dateFormat M
    axisFormat %b
    todayMarker off
    section Seasons
    Jilaal :1, 4
    Gu     :4, 7
    Post-Gu:7, 8
    Hagaa  :7, 10
    Deyr   :10, 12
```

+ It is generally assumed that AMN in Somalia peaks from the end of the Jilaal season to the end of the Hagaa season??
:::

## The regional literature {.smaller}

A few studies where conducted in the region using SMART survey data: 

::: {.incremental}
+ 2 studies using 15 years of SMART survey data **in African drylands** (excluding Somalia) underscored the complex relationship between AMN and acute food insecurity: 
  - AMN and food security seasons do not always align across seasons or livelihoods. 
  - Identified two AMN seasonal peaks: **April-May** and **August-September**  

+ Somalia specific analysis used 8 years FSNAU and ACF SMART surveys (2014-2021):  

  - Seasonal peak 1: **May/June (the beginning of Gu)** 
  - Seasonal peak 2: **November (mid-Deyr)** with a decline during the lean season.
  - There were variations in the patterns according to the livelihood system (pastoralist, agropastoralist and farming)

:::

## What does this analysis add? {.smaller} 

This analysis examines six years of severe acute malnutrition (SAM) admission data from treatment programs in Somalia to explore patterns in admissions, focusing on **trends and seasonality** over time.  

### Objectives  

  + **Identify and describe trend patterns**: a long-term direction or movement in the admissions that persists across the series. This represents underlying patterns in the admissions after removing short-term fluctuations. 

  + **Identify and describe the seasonal patterns**: recurring patterns that occur in a fixed and specific period every year. In this study, this refers to monthly recurring patterns over the years.

# Methods
## Data source {.smaller} 
[Somalia Nutrition Cluster's](https://response.reliefweb.int/somalia/nutrition) six years SAM admissions. 

  - The data is reported monthly at district level.

## Data wrangling {.smaller} 
```{mermaid}
%%| label: wrangling-workflow
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    A[(Dataset)]
    B[Exclusion of districts with consistent missingness of admissions data]
    C[Summarise admissions by **month** at national level]
    D[Summarise admissions by **month** by livelihood systems]
    E[Time series]
    F([National])
    G([By livelihood systems])

    A --> B --> C --> E
    B --> D --> E
    E --> F
    E --> G
```

::: {.callout-note}
## Excluded districts
Twelve districts were excluded for missingness: Ceel Dheere, Jalalaqsi, Jamaame, Kurtunwaarey, Sablaale, Adan Yabaal, Bu'aale, Jilib, Saakow/Salagle, Sheik, Cadale, and Xarardheere.
:::

## Analysis approach {.smaller}  
::: {.incremental}
1. Explored visually through graphs to identify patterns over time
2. It was broken down into components to examine the overall monthly trend and seasonal variation
3. The average rate of change (ARC) ^[The ARC describes the average or typical rate of change (in the admissions of SAM cases in this case) over a determined time interval of interest in the time series.] was calculated to better understand the trend  

Data was split into livelihood systems.
:::

# Results  
National
```{r}
#| label: source-script
source("script.R")
```

## A glimpse on the evolution of admissions over time {.smaller}
### National 
```{r}
#| label: fig-timeplot-national
#| fig-width: 12
#| fig-height: 4.3
tsplot_national
```

## A glimpse on the seasonal variation {.smaller}
```{r}
#| label: fig-seasonal-plot-national
#| fig-width: 13
#| fig-height: 7
ssnplot_national
```

## A glimpse on the seasonal variation {.smaller}
:::: {.columns}

::: {.column width="50%" style="text-align: center;"}
Before 2022

```{r}
#| label: fig-seasonal-plot-b2022
#| fig-height: 8
ssplot_national_b2022
```
:::

::: {.column width="50%" style="text-align: center;"}
As of 2022

```{r}
#| label: fig-seasonal-plot-a2022
#| fig-height: 8
ssplot_national_a2022
```
:::

::::

# Results  
By Livelihood Systems

## Number of districts per livelihood system

| Livelihood system | Number of districts |
| :-- | :-: |
| Pastoral | 34 |
| Agropastoral | 19 |
| Riverine | 5 |
| Urban/IDPs | 5 |  

: Number of districts in each livelihood system {#tbl-lsystem-districts}

## Evolution of admissions over time {.smaller}

```{r}
#| label: fig-timeplot-lsystem
#| fig-width: 13
#| fig-height: 7
tsplot_lsysttem
```

# Results
The time series components: **National level** 

## The components {.smaller}
```{r}
#| label: fig-cmpnts-national
#| fig-width: 13
#| fig-height: 7
cmpnts_plot_national
```

## Trend {.smaller}
```{r}
#| label: fig-piecewise-slopes
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan-Jun 2019"
#|   - "Jul-Oct 2019"
#|   - "Nov 2019-Feb 2020"
#|   - "Mar 2020-Jan 2021"
#|   - "Feb 2021-Mar 2023"
#|   - "Apr 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_beforeJul2019
slope_aug2019_oct2019
slope_nov2019_feb2020
slope_mar2020_jan2021
slope_feb2021_mar2023
slope_april2023_dec2024
```

## The average rate of change (ARC) of the trend {.smaller}  

| Slope | Time interval | Admissions~i~ | Admissions~f~ | ARC |  
| :---: | :---: | :---: | :---: | :---: |
| (a) | `r arc_beforeJul2019[[5]]` | `r format(round(arc_beforeJul2019[[1]]),F,big.mark=",")` | `r format(round(arc_beforeJul2019[[3]]),F,big.mark=",")` | `r round(arc_beforeJul2019[[6]])` |  
| (b) | `r arc_aug2019_oct2019[[5]]` | `r format(round(arc_aug2019_oct2019[[1]]),F, big.mark=",")` | `r format(round(arc_aug2019_oct2019[[3]]),F,big.mark=",")` | `r round(arc_aug2019_oct2019[[6]])` | 
| (c) | `r arc_nov2019_feb2020[[5]]` | `r format(round(arc_nov2019_feb2020[[1]]),F,big.mark=",")` | `r format(round(arc_nov2019_feb2020[[3]]),F,big.mark=",")` | `r round(arc_nov2019_feb2020[[6]])` | 
| (d) | `r arc_mar2020_jan2021[[5]]` | `r format(round(arc_mar2020_jan2021[[1]]),F,big.mark=",")` | `r format(round(arc_mar2020_jan2021[[3]]),F,big.mark=",")` | `r round(arc_mar2020_jan2021[[6]])` | 
| (e) | `r arc_feb2021_mar2023[[5]]` | `r format(round(arc_feb2021_mar2023[[1]]),F,big.mark=",")` | `r format(round(arc_feb2021_mar2023[[3]]),F, big.mark=",")` | `r format(round(arc_feb2021_mar2023[[6]]),big.mark=",")` | 
| (f) | `r arc_april2023_dec2024[[5]]` | `r format(round(arc_april2023_dec2024[[1]]),F, big.mark=",")` | `r format(round(arc_april2023_dec2024[[3]]),F,big.mark=",")` | `r format(round(arc_april2023_dec2024[[6]]),big.mark=",")` | 

: The average rate of change of admissions by month in the time-intervals {#tbl-arc-national}

## The seasonal variation {.smaller}
```{r}
#| label: fig-seasonal-pattern-national
#| fig-width: 13
#| fig-height: 7
seasonal_cmpnt_national
```

## Seasonal variation before and as of 2022 {.smaller}
:::: {.columns}

::: {.column width="50%" style="text-align: center;"}
```{r}
#| label: fig-seasonal-pattern-natioanal-b2022
#| fig-height: 8
seasonal_cmpnt_national_b2022
```
:::

::: {.column width="50%" style="text-align: center;"}
```{r}
#| label: fig-seasonal-pattern-natioanal-a2022
#| fig-height: 8
seasonal_cmpnt_national_a2022
```
:::

::::

# Results
The time series components: **By Livelihood Systems**  

  + _Pastoral_  

## The components {.smaller}
```{r}
#| label: fig-cmpnts-pastoral
#| fig-width: 13
#| fig-height: 7
cmpnts_plot_pastoral
```

## Trend {.smaller}
```{r}
#| label: fig-piecewise-slopes-pastoral
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan-May 2019"
#|   - "Jun-Sep 2019"
#|   - "Oct 2019-Aug 2020"
#|   - "Sep 2020-Apr 2021"
#|   - "May 2021-Dec 2022"
#|   - "Jan 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_pastoral_beforemay2019
slope_pastoral_jun_sep2019
slope_pastoral_oct2019_aug2020
slope_pastoral_sep2020_apr2021
slope_pastoral_may2021_dec2022
slope_pastoral_jan2023_dec2024
```

## Seasonal variation {.smaller}
```{r}
#| label: fig-seasonal-pattern-pastoral
#| fig-width: 13
#| fig-height: 7
seasonal_cmpnt_pastoral
```

# Results
The time series components: **By Livelihood Systems**  

  + _Agropastoral_  

## The components {.smaller}
```{r}
#| label: fig-cmpnts-agropastoral
#| fig-width: 13
#| fig-height: 7
cmpnts_plot_agropastoral
```

## Trend {.smaller}
```{r}
#| label: fig-piecewise-slopes-agropastoral
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan 2019-Jan 2020"
#|   - "Feb 2020-Jan 2021"
#|   - "Feb 2021-Feb 2023"
#|   - "Mar 2023-Dec 2024"
#| layout-ncol: 2
#| layout-nrow: 2

slope_agropastoral_beforefeb2020
slope_agropastoral_feb2020_jan2021
slope_agropastoral_feb2021_feb2023
slope_agropastoral_mar2023_dec2024
```

## Seasonal variation {.smaller}
```{r}
#| label: fig-seasonal-pattern-agropastoral
#| fig-width: 13
#| fig-height: 7
seasonal_cmpnt_agropastoral
```

# Results
The time series components: **By Livelihood Systems**  

  + _Riverine_  

## The components {.smaller}
```{r}
#| label: fig-cmpnts-riverine
#| fig-width: 13
#| fig-height: 7
cmpnts_plot_riverine
```

## Trend {.smaller}
```{r}
#| label: fig-piecewise-slopes-riverine
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan 2019-May 2020"
#|   - "Jun 2019-Nov 2020"
#|   - "Dec 2020-Apr 2021"
#|   - "May 2021-Sep 2021"
#|   - "Oct 2021-Aug 2023"
#|   - "Sep 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_riverine_beforemay2020
slope_riverine_jun_nov2020
slope_riverine_dec2020_apr2021
slope_riverine_may2021_sep2021
slope_riverine_oct2021_aug2023
slope_riverine_sep2023_dec2024
```

## Seasonal variation {.smaller}
```{r}
#| label: fig-seasonal-pattern-riverine
#| fig-width: 13
#| fig-height: 7
seasonal_cmpnt_riverine
```

# Results
The time series components: **By Livelihood Systems**  

  + _Urban/IDPs_  

## The components {.smaller}
```{r}
#| label: fig-cmpnts-urban-idps
#| fig-width: 13
#| fig-height: 7
cmpnts_plot_urban_idps
```

## Trend {.smaller}
```{r}
1+1
```

## Seasonal variation {.smaller}
```{r}
#| label: fig-seasonal-pattern-urban-idps
#| fig-width: 13
#| fig-height: 7
seasonal_cmpnt_urban_idps
```

# Discussion  

{{< lorem 50 >}}