---
title: "Feature insights into Somalia’s severe acute malnutrition admissions"
subtitle: "A time series analysis spanning from January 2019 to November 2024"
author: "Tomás Zaba"
date: today
---

## Context  
Seasonality significantly impacts the underlying and immediate factors contributing to acute malnutrition (AMN). It refers to recurring patterns within a year [@tsR], such as specific months or quarters when AMN cases rise or fall.  

Understanding the seasonal variation of AMN is crucial for evidence-based decision-making. However, despite a general awareness of its seasonality, limited evidence exists to substantiate it. Consequently, stakeholders often rely on assumptions of a seasonal peak during the food security lean season [@africadry]. To account for this, nutrition surveys are routinely timed to coincide with the lean season each year.

The limited understanding of AMN’s seasonal variation is attributed to two key challenges: (i) the availability of robust analytical methods and (ii) access to primary data [@africadry]. These challenges, however, can be mitigated by leveraging high-quality secondary data, such as survey datasets or treatment program admissions. Studies using 15 years of SMART survey data from African drylands have underscored the complex relationship between AMN and acute food insecurity, revealing that they do not always align across seasons or livelihoods [@africadry; @anastasia]. Notably, this research identified two AMN seasonal peaks: April-May and August-September [@africadry].  

In Somalia, the seasonality of AMN is also generaly associated with that of the food security seasonality. There are five defined food security seasons as described by @ENN:  

```{mermaid}
%%| label: seasonal-calendar
%%| fig-cap: "Source: author's adaptation"
%%| fig-cap-location: bottom
gantt
    title  Somalia Food Security Seasonal Calendar
    dateFormat M
    axisFormat %b
    todayMarker off
    section Seasons
    Jilaal :1, 4
    Gu     :4, 7
    Post-Gu:7, 8
    Hagaa  :7, 10
    Deyr   :10, 12
```

  + **Jilaal**: the most extended dry season, is characterized by livestock migration and decreased milk production in the pastoralist livelihoods. 
  + **Gu**: characterized by heavy rainfall, it constitutes Somalia's 70% of the crop production season, and livestock production depends upon this season.
  + **Post-Gu**: characterized by increase pasture and water availability for the livestock and increased availability of milk, fruits, vegetables. This is as well the peak season of a liquid butter made of milk know as *ghee*.  
  + **Hagaa**: a dry season, characterized by light rainfalls.  
  + **Deyr**: a rainy season, however localized and less reliable.

It is generally assumed that AMN in Somalia peaks from the end of the Jilaal season to the end of the Hagaa season. An analysis by @ENN, based on SMART surveys conducted by the Food Security and Nutrition Analysis Unit (FSNAU) and partners, with data collected between the Gu and Deyr seasons from 2014 to 2021, identified two distinct seasonal peaks: **May/June (the beginning of Gu)** and **November (mid-Deyr)**, with a decline during the lean season. The study also highlighted variations in these patterns based on livelihood zones, including pastoralist, agro-pastoralist, and riverine/farming communities:

  + Among pastoralist communities, the peak in AMN was observed in June (Gu). This does not align with the food security lean season, as pastoral consumption typically improves starting in April when animals give birth, coinciding with the onset of rainfall. During this period, households gain access to milk and meat and income from the sale of animals. A second peak was noted in November, a period when food insecurity is generally expected to be minimal following the Gu harvest [@ENN].

  + In agropastoralist and farming areas, the seasonality followed the peaks described above [@ENN].

### What does this analysis add?
  This analysis examines six years of severe acute malnutrition (SAM) admission data from treatment programs in Somalia to explore patterns in admissions, focusing on **trends and seasonality** over time.

#### Objectives
  + **Identify and describe trend patterns**: a long-term direction or movement in the admissions that persists across the series. This represents underlying patterns in the admissions after removing short-term fluctuations. 

  + **Identify and describe the seasonality patterns**: recurring patterns that occur in a fixed and specific period every year. 

  + **Quantify the seasonal influences on the admissions**: the extent to which different times of the year impact the number of SAM cases admitted into the treatment program

{{< pagebreak >}}

## Methods
### Data source
The analysis used the [Somalia Nutrition Cluster's](https://response.reliefweb.int/somalia/nutrition) six years of countrywide data on SAM admission into AMN treatment program. The data is reported monthly at district level. 

Initially, the analysis was intended to use data on global acute malnutrition data. However, based on the advice from the Nutrition Cluster advised, moderate acute malnutrition (MAM) was excluded. Unlike SAM data, MAM data is affected by frequent stockouts of read-to-use supplementary food (RUSF), leading to unstable variance. Including this data could result in misleading findings.

### Data wrangling {#sec-data-wrangling} 
It consisted in excluding districts with missing admissions from January to December; then by aggregating the sum of the total number of children admitted by month at national level and by livelihood systems (Pastoral, Agropastoral, Riverine and Urban/IDPs).
```{mermaid}
%%| label: wrangling-workflow
%%| fig-cap: Data wrangling workflow
flowchart LR
    A[(Dataset)]
    B[Exclusion of districts with consistent missingness of admissions data]
    C[Summarise admissions by **month** at national level]
    D[Summarise admissions by **month** by livelihood systems]
    E[Time series]
    F([National])
    G([By Livelihood systems])

    A --> B --> C --> E
    B --> D --> E
    E --> F
    E --> G
```

::: {.callout-note}
Eight districts were excluded for missingness: Ceel Dheere, Jalalaqsi, Jamaame, Kurtunwaarey Sablaale, Adan Yabaal, Bu'aale, Jilib, Saakow/Salagle, Sheik, Cadale, and Xarardheere.
:::

### Analysis approach  
The data was first explored visually through graphs to identify patterns over time. It was then broken down into components to examine the overall trend and seasonal variations. The average rate of change was calculated to better understand the trend. To account for the different livelihood systems within Somalia, the data was split into four separate time series, each representing a different livelihood system.

A statistical model was used to measure the impact of the overall trend and seasonal changes on the number of admissions.

The analysis was performed using standard statistical tools, which are available in the project’s repository for reference.

{{< pagebreak >}}

## Results  

Results will first be presented and described for data grouped at national level, and then followed by the livelihood systems.
```{r}
#| label: source-script
source("script.R")
```

### Overview of how admissions changed over time with data grouped at national level {#sec-overview-national}
Figure shows a display of the total number of SAM cases admitted into treatment program in Somalia over six years, starting from January 2019 to December 2024, on a monthly continuos scale. It shows a changing trend over time, with remarkable rise as of 2022, that was preceeded by a relatively stable trend, followed by a consistent decline as of 2023. On the other hand, it also shows an irregular seasonal pattern (the ups and downs on the figure), with the strenghts and the intensity of the peaks (amplitude) chaning over time.  

```{r}
#| label: fig-timeplot-national
tsplot_national
```

While this figure gives a glimpse of the underlying pattern of the trend and the seasonal variation in the admissions of SAM cases over time, these components are further explored and described in the following sections. 

#### An overview of the seasonal variation  
@fig-seasonal-plot-national displays the evolution of the SAM admissions across months(x-axis) by year (y-axis). Each line represents a year along the months. 

From this figure, it is starting to be seen, with some clarity, the seasonal variation. Two patterns can be spotted from the figure: (i) before 2022, the seasonal variation seemed to be regular each year, with a relatively shallow amplitude between each lines. (ii) as of 2022, both the amplitude between lines, as well as 

::: {.callout-note}
2022 was the year when Somalia was hit by a severe drought, and with that, most areas had been classified under the Risk of Famine, and two districts projected to be in Famine situation in 2022 [@ipcmanual].
:::
```{r}
#| label: fig-seasonal-plot-national
ssnplot_national
```

The patterns in the above figure can further be visualized below with the data split into admissions before 2022 and as of 2022.

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-plot-b2022
ssplot_national_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-plot-a2022
ssplot_national_a2022
```
:::

### Subseries  
The figure below is a plot of the evolution of the admissions in each month over years. It helps to visualize the patterns of the admissions in a given month as the years pass. The blue horizontal line represents the average number of admissions each month over years. 

Two patterns can be spotted from the figure: (i), the admissions are on average high in January every year, then it tends to decline until April, and then it rises in May-June, then followed by a slight decline from August to October, and then starts to rise again as of November. (ii) as of 2022, the admissions rised remarkably in every months, and then the decline in the trend is also natable in every months, with some months showing a strong decline than the others.
```{r}
#| label: fig-subseries-national
sbsplot_national
```

The data can be visualized with splits before and as of 2022 as well. The underlying patterns gleaned are in accordance with those described in @fig-seasonal-plot-b2022 (for before 2022) and @fig-seasonal-plot-a2022 (for as of 2022).  

::: {.callout-tip}
Focus on the blue horizontal lines. Read the patterns they depict, and compare with those from the referenced figures above.
:::

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-subseries-plot-b2022
sbsplot_national_b2022
```

## As of 2022
```{r}
#| label: fig-subseries-plot-a2022
sbsplot_national_a2022
```
:::

### Overview of how admissions changed by over time by livelihood system
#### Distribution of the number of districts per livelihood system

Majority of the districts in Somalia are pastoral (34 districts), followed by agropastoral (19 district) @tbl-lsystem-districts. The table counts districts after excluding those with missing values, as described in @sec-data-wrangling.

| Livelihood system | Number of districts |
| :-- | :-: |
| Pastoral | 34 |
| Agropastoral | 19 |
| Riverine | 5 |
| Urban/IDPs | 5 |  

: Number of districts in each livelihood system {#tbl-lsystem-districts}

#### How admissions changed over time 

Similar to @fig-timeplot-national, figure 7 shows the evolution of the SAM admissions over time, however this time with the data split into the four livelihood systems. All in all, the admissions in each livelihood system seems to follow the same pattern as that described in @sec-overview-national.

```{r}
#| label: fig-timeplot-lsystem
tsplot_lsysttem
```

{{< pagebreak >}}

### Identifying the trends and seasonal patterns in admissions over time  
A time series can exhibit a variety of underlying patterns that may not be easly visible from a regular time plot. These can be gleaned by spliting the time series into its constituent components and clearly see them. The following sections displays the gleaned underlying patterns of the admissions data. The description starts with national level data, and then by livelihood system. 

#### At National level  
In @fig-cmpnts-national, the first graph (top-down) is the whole data (the admissions). This is the same graph shown in @fig-timeplot-national. The second graph is the **trend pattern alone** over time. The thrid graph is **seasonal pattern alone** over time after removing the trend. The fourth graph represents any other variation in the admissions that are not caused by the trend or the seasonality.  

```{r}
#| label: fig-cmpnts-national
cmpnts_plot_national
```

##### The trend component 

All in all, as it was mentioned in the overview of the admissions (@sec-overview-national), in terms of **shape**, the trend pattern is non-linear; in terms of **stability**, the trend pattern is a changing trend that evolves over time (not stable). In terms of the **direction**, the trend shows that the admissions rose remarkably as of the last quarter of 2021, and started to decline in 2023.  

###### The ARC of the admissions in the times the slope of the admissions changed remarkably

```{r}
#| label: fig-piecewise-slopes
#| fig-cap: "Trend line in pieces"
#| fig-subcap: 
#|   - "Slope Jan-Jun 2019"
#|   - "Slope Jul-Oct 2019"
#|   - "Slope Nov 2019-Feb 2020"
#|   - "Slope Mar 2020-Jan 2021"
#|   - "Slope Feb 2021-Mar 2023"
#|   - "Slope Apr 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_beforeJul2019

slope_aug2019_oct2019

slope_nov2019_feb2020

slope_mar2020_jan2021

slope_feb2021_mar2023

slope_april2023_dec2024
```

Considering the different remarkable fluctuations observed in the pattern at specific time-period, an average rate of change (ARC) [^1] [@arc] was calculated for each pieces of the above slopes. Results are presented in @tbl-arc-national.

::: {.callout-note}
## How to read the table 2 in conjuntion with figure 10 

**Example 1: increase**  
In slope (a), the trend starts in January 2019 with an average admission of `r format(round(arc_beforeJul2019[[1]]),F,big.mark=",")` cases. It increased to an average admission of `r format(round(arc_beforeJul2019[[3]]),F,big.mark=",")` cases by June of the same year. Between this period of `r arc_beforeJul2019[[5]]` months, the admissions increased by an average of `r round(arc_beforeJul2019[[6]])` cases each month.

**Example 2: decrease**  
In slope (e), the trend starts in April 2023, with an average of admission of `r format(round(arc_april2023_dec2024[[1]]),F, big.mark=",")` cases. It decreased to an average admission of `r format(round(arc_april2023_dec2024[[3]]),F,big.mark=",")` cases by December 2024. During this period of `r arc_april2023_dec2024[[5]]` months, the admissions decreased by an average of `r format(round(arc_april2023_dec2024[[6]]),big.mark=",")` cases each month. 
:::

[^1]: The ARC descrobes the average or typical rate of change (in the admissions of SAM cases in this case) over a determined time interval of interest in the time series. 

| Slope | Time interval | Admissions~i~ | Admissions~f~ | ARC |  
| :---: | :---: | :---: | :---: | :---: |
| (a) | `r arc_beforeJul2019[[5]]` | `r format(round(arc_beforeJul2019[[1]]),F,big.mark=",")` | `r format(round(arc_beforeJul2019[[3]]),F,big.mark=",")` | `r round(arc_beforeJul2019[[6]])` |  
| (b) | `r arc_aug2019_oct2019[[5]]` | `r format(round(arc_aug2019_oct2019[[1]]),F, big.mark=",")` | `r format(round(arc_aug2019_oct2019[[3]]),F,big.mark=",")` | `r round(arc_aug2019_oct2019[[6]])` | 
| (c) | `r arc_nov2019_feb2020[[5]]` | `r format(round(arc_nov2019_feb2020[[1]]),F,big.mark=",")` | `r format(round(arc_nov2019_feb2020[[3]]),F,big.mark=",")` | `r round(arc_nov2019_feb2020[[6]])` | 
| (d) | `r arc_mar2020_jan2021[[5]]` | `r format(round(arc_mar2020_jan2021[[1]]),F,big.mark=",")` | `r format(round(arc_mar2020_jan2021[[3]]),F,big.mark=",")` | `r round(arc_mar2020_jan2021[[6]])` | 
| (e) | `r arc_feb2021_mar2023[[5]]` | `r format(round(arc_feb2021_mar2023[[1]]),F,big.mark=",")` | `r format(round(arc_feb2021_mar2023[[3]]),F, big.mark=",")` | `r format(round(arc_feb2021_mar2023[[6]]),big.mark=",")` | 
| (f) | `r arc_april2023_dec2024[[5]]` | `r format(round(arc_april2023_dec2024[[1]]),F, big.mark=",")` | `r format(round(arc_april2023_dec2024[[3]]),F,big.mark=",")` | `r format(round(arc_april2023_dec2024[[6]]),big.mark=",")` | 

: The average rate of change in the admissions by month in the time-intervals represented in figure 10 {#tbl-arc-national}

##### Seasonal pattern   
```{r}
#| label: fig-seasonal-pattern-national
seasonal_cmpnt_national
```



::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-natioanal-b2022
seasonal_cmpnt_national_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-natioanal-a2022
seasonal_cmpnt_national_a2022
```
:::

#### By Livelihood systems  
##### Pastoral

```{r}
#| label: fig-cmpnts-pastoral
cmpnts_plot_pastoral
```

###### Seasonal pattern   
```{r}
#| label: fig-seasonal-pattern-pastoral
seasonal_cmpnt_pastoral
```

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-pastoral-b2022
seasonal_cmpnt_pastoral_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-pastoral-a2022
seasonal_cmpnt_pastoral_a2022
```
:::

##### Agropastoral

```{r}
#| label: fig-cmpnts-agropastoral
cmpnts_plot_agropastoral
```

###### Seasonal pattern   
```{r}
#| label: fig-seasonal-pattern-agropastoral
seasonal_cmpnt_agropastoral
```

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-agropastoral-b2022
seasonal_cmpnt_agropastoral_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-agropastoral-a2022
seasonal_cmpnt_agropastoral_a2022
```
:::

##### Riverine

```{r}
#| label: fig-cmpnts-riverine
cmpnts_plot_riverine
```

###### Seasonal pattern   
```{r}
#| label: fig-seasonal-pattern-riverine
seasonal_cmpnt_riverine
```

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-riverine-b2022
seasonal_cmpnt_riverine_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-riverine-a2022
seasonal_cmpnt_riverine_a2022
```
:::

##### Urban/IDPs

```{r}
#| label: fig-cmpnts-urban-idps
cmpnts_plot_urban_idps
```

###### Seasonal pattern   
```{r}
#| label: fig-seasonal-pattern-urban-idps
seasonal_cmpnt_urban_idps
```

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-urban-idps-b2022
seasonal_cmpnt_urban_idps_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-urban-idps-a2022
seasonal_cmpnt_urban_idps_a2022
```
:::

{{< pagebreak >}}