---
title: "Feature insights into Somalia’s severe acute malnutrition admissions"
subtitle: "A time series analysis spanning from January 2019 to November 2024"
author: "Tomás Zaba"
date: today
---

## Context  
Seasonality significantly impacts the underlying and immediate factors contributing to acute malnutrition (AMN). It refers to recurring patterns within a year [@tsR], such as specific months or quarters when AMN cases rise or fall.  

Understanding the seasonal variation of AMN is crucial for evidence-based decision-making. Nonetheless, despite a general awareness of its seasonality, limited evidence exists to substantiate it. Consequently, stakeholders often rely on assumptions of an AMN seasonal peak during the food security lean season [@africadry] and, to account for this, nutrition surveys are routinely timed to coincide with the lean season each year.

The limited understanding of AMN’s seasonal variation is attributed to two key challenges: (i) the availability of robust analytical methods and (ii) access to primary data [@africadry]. The second challenge, can be mitigated by leveraging high-quality secondary data, such as survey datasets or treatment program admissions. Studies using 15 years of SMART survey data from African drylands have underscored the complex relationship between AMN and acute food insecurity, revealing that they do not always align across seasons or livelihoods [@africadry; @anastasia]. Notably, this research identified two AMN seasonal peaks: April-May and August-September [@africadry].  

In Somalia, the seasonality of AMN is generaly associated with the seasonality of food security as well. There are five defined food security seasons as described by @ENN:  

```{mermaid}
%%| label: seasonal-calendar
%%| fig-cap: "Source: author's adaptation"
%%| fig-cap-location: bottom
gantt
    title  Somalia Food Security Seasonal Calendar
    dateFormat M
    axisFormat %b
    todayMarker off
    section Seasons
    Jilaal :1, 4
    Gu     :4, 7
    Post-Gu:7, 8
    Hagaa  :7, 10
    Deyr   :10, 12
```

  + **Jilaal**: the most extended dry season, is characterized by livestock migration and decreased milk production in the pastoralist livelihoods. 
  + **Gu**: characterized by heavy rainfall, it constitutes Somalia's 70% of the crop production season, and livestock production depends upon this season.
  + **Post-Gu**: characterized by increase pasture and water availability for the livestock and increased availability of milk, fruits, vegetables. This is as well the peak season of a liquid butter made of milk know as *ghee*.  
  + **Hagaa**: a dry season, characterized by light rainfalls.  
  + **Deyr**: a rainy season, however localized and less reliable.

It is generally assumed that AMN in Somalia peaks from the end of the Jilaal season to the end of the Hagaa season??. An analysis by @ENN, based on SMART surveys conducted by the Food Security and Nutrition Analysis Unit (FSNAU) and partners, with data collected between the Gu and Deyr seasons from 2014 to 2021, identified two distinct seasonal peaks: **May/June (the beginning of Gu)** and **November (mid-Deyr)**, with a decline during the lean season. The study also highlighted variations in these patterns based on livelihood zones, including pastoralist, agro-pastoralist, and riverine/farming communities:

  + Among pastoralist communities, the peak in AMN was observed in June (Gu), which did not align with the food security lean season, as pastoral consumption typically improves starting in April when animals give birth, coinciding with the onset of rainfall; during this period, households gain access to milk and meat and income from the sale of animals [@ENN]. Additionally, a second peak was noted in November, a period when food insecurity is generally expected to be minimal following the Gu harvest [@ENN].

  + In agropastoralist and farming areas, the seasonality followed the peaks described above [@ENN].

### What does this analysis add?
  This analysis examines six years of severe acute malnutrition (SAM) admission data from treatment programs in Somalia to explore patterns in admissions, focusing on **trends and seasonality** over time.

#### Objectives
  + **Identify and describe trend patterns**: a long-term direction or movement in the admissions that persists across the series. This represents underlying patterns in the admissions after removing short-term fluctuations. 

  + **Identify and describe the seasonal patterns**: recurring patterns that occur in a fixed and specific period every year. In this study, this refers to monthly recurring patterns over the years.

{{< pagebreak >}}

## Methods
### Data source
The analysis used the [Somalia Nutrition Cluster's](https://response.reliefweb.int/somalia/nutrition) six years of countrywide data on SAM admission into AMN treatment program. The data is reported monthly at district level. 

The initial analysis aimed to use data on global acute malnutrition. However, data on moderate acute malnutrition (MAM) was excluded due to unstable variance caused by frequenty stockouts of read-to-use supplementary food (RUSF). Including MAM data in the analysis could have led to misleading findings.

### Data wrangling {#sec-data-wrangling} 
It consisted in excluding districts with missing admissions from January to December; then by aggregating the sum of the total number of children admitted by month at national level, and by livelihood systems (Pastoral, Agropastoral, Riverine and Urban/IDPs).
```{mermaid}
%%| label: wrangling-workflow
%%| fig-cap: Data wrangling workflow
flowchart LR
    A[(Dataset)]
    B[Exclusion of districts with consistent missingness of admissions data]
    C[Summarise admissions by **month** at national level]
    D[Summarise admissions by **month** by livelihood systems]
    E[Time series]
    F([National])
    G([By livelihood systems])

    A --> B --> C --> E
    B --> D --> E
    E --> F
    E --> G
```

::: {.callout-note}
Twelve districts were excluded for missingness: Ceel Dheere, Jalalaqsi, Jamaame, Kurtunwaarey, Sablaale, Adan Yabaal, Bu'aale, Jilib, Saakow/Salagle, Sheik, Cadale, and Xarardheere.
:::

### Analysis approach  
The data was first explored visually through graphs to identify patterns over time. Then, it was broken down into components to examine the overall monthly trend and seasonal variation. The average rate of change (ARC) [^1] was calculated to better understand the trend. To account for the different livelihood systems within Somalia, the data was split into four separate time series, each representing a different livelihood system.

The analysis was performed using standard statistical tools, which are available in the project’s repository for reference. 

[^1]: The ARC describes the average or typical rate of change (in the admissions of SAM cases in this case) over a determined time interval of interest in the time series. 

{{< pagebreak >}}

## Results  

Results will first be presented and described for data grouped at national level, and then by livelihood systems.
```{r}
#| label: source-script
source("script.R")
```

### A glimpse on the evolution of admissions over time {#sec-glimpse-national}  
#### At National-level  
The figure below displays the total number of SAM cases admitted into treatment programs in Somalia over six years, from January 2019 to December 2024, on a continuous monthly scale. It highlights a shifting trend over time: a relatively stable period before 2022, followed by a significant rise, and then a consistent decline beginning in 2023. Additionally, the figure reveals the existence of a seasonal pattern (the ups and downs), however it seems to vary over time.

```{r}
#| label: fig-timeplot-national
tsplot_national
```

##### A glimpse on the seasonal variation  
@fig-seasonal-plot-national displays the evolution of admissions across months(x-axis) by year (y-axis). Each line represents a year along the months. 

From this figure, it is starting to be seen, with some clarity, the seasonal variation. Two patterns can be spotted from the figure:  

  1. Before 2022, the seasonal variation seemed to be regular each year, with a relatively shallow amplitude between each lines.  
  2. As of 2022, both the amplitude between lines rose. 

```{r}
#| label: fig-seasonal-plot-national
ssnplot_national
```

The patterns in the above figure can further be visualized below with the data split into admissions before 2022 and as of 2022.

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-plot-b2022
ssplot_national_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-plot-a2022
ssplot_national_a2022
```
:::

#### By livelihood systems  
The majority of the districts in Somalia are pastoral (34 districts), followed by agropastoral (19 districts) @tbl-lsystem-districts. The table counts districts after excluding those with missing values, as described in @sec-data-wrangling.

| Livelihood system | Number of districts |
| :-- | :-: |
| Pastoral | 34 |
| Agropastoral | 19 |
| Riverine | 5 |
| Urban/IDPs | 5 |  

: Number of districts in each livelihood system {#tbl-lsystem-districts}

All in all, the admissions in each livelihood system seem to follow the same pattern described in @sec-glimpse-national.

```{r}
#| label: fig-timeplot-lsystem
tsplot_lsysttem
```

::: {.callout-note}
While this figure gives a glimpse of the underlying pattern of the trend and the seasonal variation of admissions over time, these components are further analyzed and described in detail in the following sections.
:::
{{< pagebreak >}}

### Unveiling the trend and seasonal patterns in admissions  
A time series can exhibit a variety of underlying patterns that may not be easliy visible from a regular time plot (attempt made in above sections, however without a clear visual). By decomposing the time series into its constituent components, these hidden patterns can be brought to light. The following sections reveal the extracted patterns from the admissions data, beginning with the national-level analysis, followed by insights at the livelihood system. 

#### At National-level  
In the figure below, the first graph (top-down) is the whole data (the admissions). This is the same graph shown in @fig-timeplot-national. The second graph is the **trend pattern alone** over time. The third graph is **seasonal pattern alone** over time after removing the trend. The fourth graph represents any other variation in the admissions that is not caused by the trend or by the seasonality.  

```{r}
#| label: fig-cmpnts-national
cmpnts_plot_national
```

##### Analyzing the trend component 

All in all, as highlighted in the overview of admissions (@sec-glimpse-national), the trend pattern demonstrates distinct characteristics. In terms of **shape**, it is non-linear. Regarding **stability**, the trend evolves over time, indicating a dynamic rather than stable pattern. In terms of **direction**, it exhibits an notable rise in admissions starting in the last quarter of 2021, followed by a decline that started to decline in 2023.  

The figure below illustrates the trend line split into the times when there was a significant change in the direction of the trend over time.

```{r}
#| label: fig-piecewise-slopes
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan-Jun 2019"
#|   - "Jul-Oct 2019"
#|   - "Nov 2019-Feb 2020"
#|   - "Mar 2020-Jan 2021"
#|   - "Feb 2021-Mar 2023"
#|   - "Apr 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_beforeJul2019
slope_aug2019_oct2019
slope_nov2019_feb2020
slope_mar2020_jan2021
slope_feb2021_mar2023
slope_april2023_dec2024
```

The ARC [@arc] was calculated for each slope. The results are summarized in @tbl-arc-national.

::: {#nte-read-arc .callout-note}
## How to read the table 2 in conjuntion with the figure 10

**Example 1: increase**  
In slope (a), the trend starts in January 2019 with an average admission of `r format(round(arc_beforeJul2019[[1]]),F,big.mark=",")` cases. It increased to an average admission of `r format(round(arc_beforeJul2019[[3]]),F,big.mark=",")` cases by June of the same year. Between this period of `r arc_beforeJul2019[[5]]` months, the admissions increased by an average of `r round(arc_beforeJul2019[[6]])` cases each month.

**Example 2: decrease**  
In slope (e), the trend starts in April 2023, with an average of admission of `r format(round(arc_april2023_dec2024[[1]]),F, big.mark=",")` cases. It decreased to an average admission of `r format(round(arc_april2023_dec2024[[3]]),F,big.mark=",")` cases by December 2024. During this period of `r arc_april2023_dec2024[[5]]` months, the admissions decreased by an average of `r format(round(arc_april2023_dec2024[[6]]),big.mark=",")` cases each month. 
:::

| Slope | Time interval | Admissions~i~ | Admissions~f~ | ARC |  
| :---: | :---: | :---: | :---: | :---: |
| (a) | `r arc_beforeJul2019[[5]]` | `r format(round(arc_beforeJul2019[[1]]),F,big.mark=",")` | `r format(round(arc_beforeJul2019[[3]]),F,big.mark=",")` | `r round(arc_beforeJul2019[[6]])` |  
| (b) | `r arc_aug2019_oct2019[[5]]` | `r format(round(arc_aug2019_oct2019[[1]]),F, big.mark=",")` | `r format(round(arc_aug2019_oct2019[[3]]),F,big.mark=",")` | `r round(arc_aug2019_oct2019[[6]])` | 
| (c) | `r arc_nov2019_feb2020[[5]]` | `r format(round(arc_nov2019_feb2020[[1]]),F,big.mark=",")` | `r format(round(arc_nov2019_feb2020[[3]]),F,big.mark=",")` | `r round(arc_nov2019_feb2020[[6]])` | 
| (d) | `r arc_mar2020_jan2021[[5]]` | `r format(round(arc_mar2020_jan2021[[1]]),F,big.mark=",")` | `r format(round(arc_mar2020_jan2021[[3]]),F,big.mark=",")` | `r round(arc_mar2020_jan2021[[6]])` | 
| (e) | `r arc_feb2021_mar2023[[5]]` | `r format(round(arc_feb2021_mar2023[[1]]),F,big.mark=",")` | `r format(round(arc_feb2021_mar2023[[3]]),F, big.mark=",")` | `r format(round(arc_feb2021_mar2023[[6]]),big.mark=",")` | 
| (f) | `r arc_april2023_dec2024[[5]]` | `r format(round(arc_april2023_dec2024[[1]]),F, big.mark=",")` | `r format(round(arc_april2023_dec2024[[3]]),F,big.mark=",")` | `r format(round(arc_april2023_dec2024[[6]]),big.mark=",")` | 

: The average rate of change of admissions by month in the time-intervals represented in figure 10 {#tbl-arc-national}

##### Examining the seasonal component  {#sec-sp-national}  
All in all, the seasonal component of the national-level data reveals a consistent seasonal pattern in the admissions, though it is **time-varying**. The **highest peak** of admissions occurs in **May-June**, preceded by a **lower trough consistently observed in April**, which starts in February. The first peak occurs in January, following an increase in admissions starting in November.

While the patterns remain consistent over the years, the amplitude[^2] of both the peaks and troughs has varied over time. Prior to 2022, it was higher and sharper, while the trough amplitude was relatively low. Conversely, from 2022 onwards, the peak amplitude decreased, while the trough amplitude increased significantly.  

[^2]: The amplitude of a peak or trough refers to the difference between the peak or trough's value and the average trend the level. The latter is represented by zero on the y-axis of the seasonal plot. 

```{r}
#| label: fig-seasonal-pattern-national
seasonal_cmpnt_national
```

The seasonal component before and as of 2022 can be visualized in @fig-seasonal-pattern-natioanal-b2022 and in @fig-seasonal-pattern-natioanal-a2022 respectively

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-natioanal-b2022
seasonal_cmpnt_national_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-natioanal-a2022
seasonal_cmpnt_national_a2022
```
:::

#### By Livelihood systems  
##### Pastoral  

The figure below illustrates the components of admissions in the pastoral livelihood system, which are further detailed in the following sections. 

```{r}
#| label: fig-cmpnts-pastoral
cmpnts_plot_pastoral
```

###### Analyzing the trend component  
The trend’s direction has undergone six significant shifts, as illustrated below, with the ARC summarized in @tbl-arc-pastoral.
```{r}
#| label: fig-piecewise-slopes-pastoral
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan-May 2019"
#|   - "Jun-Sep 2019"
#|   - "Oct 2019-Aug 2020"
#|   - "Sep 2020-Apr 2021"
#|   - "May 2021-Dec 2022"
#|   - "Jan 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_pastoral_beforemay2019
slope_pastoral_jun_sep2019
slope_pastoral_oct2019_aug2020
slope_pastoral_sep2020_apr2021
slope_pastoral_may2021_dec2022
slope_pastoral_jan2023_dec2024
```

To read and interpret this table, refer to @nte-read-arc.

| Slope | Time interval | Admissions~i~ | Admissions~f~ | ARC |  
| :---: | :---: | :---: | :---: | :---: |
| (a) | `r arc_pastoral_beforemay2019[[5]]` | `r format(round(arc_pastoral_beforemay2019[[1]]),F,big.mark=",")` | `r format(round(arc_pastoral_beforemay2019[[3]]),F,big.mark=",")` | `r round(arc_pastoral_beforemay2019[[6]])` |  
| (b) | `r arc_pastoral_jun_sep2019[[5]]` | `r format(round(arc_pastoral_jun_sep2019[[1]]),F, big.mark=",")` | `r format(round(arc_pastoral_jun_sep2019[[3]]),F,big.mark=",")` | `r round(arc_pastoral_jun_sep2019[[6]])` | 
| (c) | `r arc_pastoral_oct2019_aug2020[[5]]` | `r format(round(arc_pastoral_oct2019_aug2020[[1]]),F,big.mark=",")` | `r format(round(arc_pastoral_oct2019_aug2020[[3]]),F,big.mark=",")` | `r round(arc_pastoral_oct2019_aug2020[[6]])` | 
| (d) | `r arc_pastoral_sep2020_apr2021[[5]]` | `r format(round(arc_pastoral_sep2020_apr2021[[1]]),F,big.mark=",")` | `r format(round(arc_pastoral_sep2020_apr2021[[3]]),F,big.mark=",")` | `r round(arc_pastoral_sep2020_apr2021[[6]])` | 
| (e) | `r arc_pastoral_may2021_dec2022[[5]]` | `r format(round(arc_pastoral_may2021_dec2022[[1]]),F,big.mark=",")` | `r format(round(arc_pastoral_may2021_dec2022[[3]]),F, big.mark=",")` | `r format(round(arc_pastoral_may2021_dec2022[[6]]),big.mark=",")` | 
| (f) | `r arc_pastoral_jan2023_dec2024[[5]]` | `r format(round(arc_pastoral_jan2023_dec2024[[1]]),F, big.mark=",")` | `r format(round(arc_pastoral_jan2023_dec2024[[3]]),F,big.mark=",")` | `r format(round(arc_pastoral_jan2023_dec2024[[6]]),big.mark=",")` | 

: The average rate of change of admissions by month in the time-intervals represented in @fig-piecewise-slopes-pastoral {#tbl-arc-pastoral}  

###### Examining the seasonal component  {#sec-sp-pastoral}   
In this livelihood system, the seasonal pattern mirrors that observed in @sec-sp-national, with a few key differences. It exhibits three distinct peaks: the highest peak in May-June, a second peak in January, and a third peak in November, with the increase beginning in September. Additionally, there is a more pronounced trough in April. Notably, the seasonal patterns underwent a shift starting in 2022.

```{r}
#| label: fig-seasonal-pattern-pastoral
seasonal_cmpnt_pastoral
```

As of 2022, the first high peak shifted to January (@fig-seasonal-pattern-pastoral-a2022). The May-June period maintained a high peak, with an amplitude similar to the November peak, which begins in October and continues through January. From January onward, there is a gradual decline, less intense compared to the sharper decline observed in the pre-2022 pattern. Regarding the trough, its amplitude significantly increased starting in 2022.

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-pastoral-b2022
seasonal_cmpnt_pastoral_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-pastoral-a2022
seasonal_cmpnt_pastoral_a2022
```
:::

##### Agropastoral  
The figure below illustrates the components of admissions in the agropastoral livelihood system, which are further detailed in the following sections. 

```{r}
#| label: fig-cmpnts-agropastoral
cmpnts_plot_agropastoral
```

###### Analyzing the trend component   
The trend’s direction has undergone four significant shifts, as illustrated below, with the ARC summarized in @tbl-arc-riverine.
```{r}
#| label: fig-piecewise-slopes-agropastoral
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan 2019-Jan 2020"
#|   - "Feb 2020-Jan 2021"
#|   - "Feb 2021-Feb 2023"
#|   - "Mar 2023-Dec 2024"
#| layout-ncol: 2
#| layout-nrow: 2

slope_agropastoral_beforefeb2020
slope_agropastoral_feb2020_jan2021
slope_agropastoral_feb2021_feb2023
slope_agropastoral_mar2023_dec2024
```


To read and interpret this table, refer to @nte-read-arc.

| Slope | Time interval | Admissions~i~ | Admissions~f~ | ARC |  
| :---: | :---: | :---: | :---: | :---: |
| (a) | `r arc_agropastoral_beforefeb2020[[5]]` | `r format(round(arc_agropastoral_beforefeb2020[[1]]),F,big.mark=",")` | `r format(round(arc_agropastoral_beforefeb2020[[3]]),F,big.mark=",")` | `r round(arc_agropastoral_beforefeb2020[[6]])` |  
| (b) | `r arc_agropastoral_feb2020_jan2021[[5]]` | `r format(round(arc_agropastoral_feb2020_jan2021[[1]]),F, big.mark=",")` | `r format(round(arc_agropastoral_feb2020_jan2021[[3]]),F,big.mark=",")` | `r round(arc_agropastoral_feb2020_jan2021[[6]])` | 
| (c) | `r arc_agropastoral_feb2021_feb2023[[5]]` | `r format(round(arc_agropastoral_feb2021_feb2023[[1]]),F,big.mark=",")` | `r format(round(arc_agropastoral_feb2021_feb2023[[3]]),F,big.mark=",")` | `r round(arc_agropastoral_feb2021_feb2023[[6]])` | 
| (d) | `r arc_agropastoral_mar2023_dec2024[[5]]` | `r format(round(arc_agropastoral_mar2023_dec2024[[1]]),F,big.mark=",")` | `r format(round(arc_agropastoral_mar2023_dec2024[[3]]),F,big.mark=",")` | `r round(arc_agropastoral_mar2023_dec2024[[6]])` | 

: The average rate of change of admissions by month in the time-intervals represented in @fig-piecewise-slopes-agropastoral {#tbl-arc-agropastoral}  

###### Examining the seasonal component   
In this livelihood system, the seasonal pattern largely mirrors those described in @sec-sp-national. The highest peak is observed in May-June, with a second peak in January, beginning its rise in December. The lowest trough is also observed in April.

```{r}
#| label: fig-seasonal-pattern-agropastoral
seasonal_cmpnt_agropastoral
```

As shown in @fig-seasonal-pattern-agropastoral-b2022, before 2022, the seasonal pattern exhibited a higher amplitude during the May-June peak. However, as of 2022, the amplitude of the May-June peak decreased, while the amplitude of the April trough significantly increased, as illustrated in @fig-seasonal-pattern-agropastoral-a2022.

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-agropastoral-b2022
seasonal_cmpnt_agropastoral_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-agropastoral-a2022
seasonal_cmpnt_agropastoral_a2022
```
:::

##### Riverine  {#sec-riverine}
The figure below illustrates the components of admissions in the riverine livelihood system, which are further detailed in the following sections

```{r}
#| label: fig-cmpnts-riverine
cmpnts_plot_riverine
```

###### Analyzing the trend component  
The trend’s direction has undergone six significant shifts, as illustrated below, with the ARC summarized in @tbl-arc-riverine
```{r}
#| label: fig-piecewise-slopes-riverine
#| fig-cap: "Periods of significant shifts in trend direction"
#| fig-subcap: 
#|   - "Jan 2019-May 2020"
#|   - "Jun 2019-Nov 2020"
#|   - "Dec 2020-Apr 2021"
#|   - "May 2021-Sep 2021"
#|   - "Oct 2021-Aug 2023"
#|   - "Sep 2023-Dec 2024"
#| layout-ncol: 3
#| layout-nrow: 2

slope_riverine_beforemay2020
slope_riverine_jun_nov2020
slope_riverine_dec2020_apr2021
slope_riverine_may2021_sep2021
slope_riverine_oct2021_aug2023
slope_riverine_sep2023_dec2024
```


To read and interpret this table, refer to @nte-read-arc.

| Slope | Time interval | Admissions~i~ | Admissions~f~ | ARC |  
| :---: | :---: | :---: | :---: | :---: |
| (a) | `r arc_riverine_beforemay2020[[5]]` | `r format(round(arc_riverine_beforemay2020[[1]]),F,big.mark=",")` | `r format(round(arc_riverine_beforemay2020[[3]]),F,big.mark=",")` | `r round(arc_riverine_beforemay2020[[6]])` |  
| (b) | `r arc_riverine_jun_nov2020[[5]]` | `r format(round(arc_riverine_jun_nov2020[[1]]),F, big.mark=",")` | `r format(round(arc_riverine_jun_nov2020[[3]]),F,big.mark=",")` | `r round(arc_riverine_jun_nov2020[[6]])` | 
| (c) | `r arc_riverine_dec2020_apr2021[[5]]` | `r format(round(arc_riverine_dec2020_apr2021[[1]]),F,big.mark=",")` | `r format(round(arc_riverine_dec2020_apr2021[[3]]),F,big.mark=",")` | `r round(arc_riverine_dec2020_apr2021[[6]])` | 
| (d) | `r arc_riverine_may2021_sep2021[[5]]` | `r format(round(arc_riverine_may2021_sep2021[[1]]),F,big.mark=",")` | `r format(round(arc_riverine_may2021_sep2021[[3]]),F,big.mark=",")` | `r round(arc_riverine_may2021_sep2021[[6]])` | 
| (e) | `r arc_riverine_oct2021_aug2023[[5]]` | `r format(round(arc_riverine_oct2021_aug2023[[1]]),F,big.mark=",")` | `r format(round(arc_riverine_oct2021_aug2023[[3]]),F,big.mark=",")` | `r round(arc_riverine_oct2021_aug2023[[6]])` | 
| (f) | `r arc_riverine_sep2023_dec2024[[5]]` | `r format(round(arc_riverine_sep2023_dec2024[[1]]),F,big.mark=",")` | `r format(round(arc_riverine_sep2023_dec2024[[3]]),F, big.mark=",")` | `r format(round(arc_riverine_sep2023_dec2024[[6]]),big.mark=",")` | 

: The average rate of change of admissions by month in the time-intervals represented in @fig-piecewise-slopes-riverine {#tbl-arc-riverine}  


###### Examining the seasonal component   
The seasonal pattern in this livelihood system demonstrates a distinctly time-varying nature, with significant changes in amplitude, as clearly shown in the “before and as of 2022” plots.

```{r}
#| label: fig-seasonal-pattern-riverine
seasonal_cmpnt_riverine
```

Before 2022, the highest peak occurred in December and the second peak in May-June. Two distinct troughs were observed, one in April and the other in November, both with nearly the same amplitude every year (@fig-seasonal-pattern-riverine-b2022).

However, as of 2022, there was a significant change in the amplitude patterns. Several peaks now appear throughout the year. Unlike the other livelihood systems, the months of May and June no longer exhibit the highest peak. Instead, the highest peak is observed in December, followed by three additional peaks of similar amplitude in January, March, and August (@fig-seasonal-pattern-riverine-a2022). Notably, in the Riverine system, and in contrast to all other areas, the month of March shows an increase in admissions.

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-riverine-b2022
seasonal_cmpnt_riverine_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-riverine-a2022
seasonal_cmpnt_riverine_a2022
```
:::

:::{.callout-note}
The Riverine and Pastoral livelihood systems are the only ones where, after the drop in admissions in February, there is an increase in March before the higher trough in April.
:::
##### Urban/IDPs  
The figure below illustrates the components of admissions in the Urban/IDPs livelihood system, which are further detailed in the following sections

```{r}
#| label: fig-cmpnts-urban-idps
cmpnts_plot_urban_idps
```

###### Examining the seasonal component   
Interestingly, in the Urban/IDP livelihood system, the seasonal pattern shows that admissions in January tend to align with the trend level. From February onwards, admissions decrease to reach a lower trough in April. Between May and June, admissions rise to their higher peak, followed by a second peak in November.

:::{.callout-important}
This is the only livelihood system where from December to April, the admissions pattern exhibits a consistent decrease every year. **Where might this be stemming from?**
:::
```{r}
#| label: fig-seasonal-pattern-urban-idps
seasonal_cmpnt_urban_idps
```

Although the pattern exhibits time-varying fluctuations, the only notable differences since 2022 are the changes observed in August and October.  

::: {.panel-tabset}
## Before 2022
```{r}
#| label: fig-seasonal-pattern-urban-idps-b2022
seasonal_cmpnt_urban_idps_b2022
```

## As of 2022
```{r}
#| label: fig-seasonal-pattern-urban-idps-a2022
seasonal_cmpnt_urban_idps_a2022
```
:::

{{< pagebreak >}}  

## Discussion  
Tackling acute malnutrition necessitates comprehensive and up-to-date information on its evolution over time and space, the times of the year when cases rise or fall, and its corresponding magnitude. This analysis aimed to provide evidence-based information on the evolution of the trend of SAM cases and its corresponding seasonal variation over the past 6 years. It was based on the use of SAM admission data as they are.  It is worth remembering that admissions are fundamentally a function of case-finding capacity and program effectiveness. Consequently, the analysis results are, also, a reflection of that. This means that if these factors are suboptimal, the observed patterns may represent false patterns, or artifacts of incomplete data. However, the data used in this analysis had an optimal (≥80%) reporting rate across catchement areas over months and years. This strengthens the reliability of the data and the confidence in the patterns gleaned, suggesting that they likely reflect the genuine trends and seasonal variation in SAM admissions rather than being result of incomplete data.

All in all, the analysis reveals that the **magnitude of children admitted into treatment programs has increased remarkably** as of the third quarter of 2021, until in 2023 (`r arc_feb2021_mar2023[[5]]` months) when it started to exhibit a downward trend (@tbl-arc-national). The average number of cases admitted during this period increased by `r format(round(arc_feb2021_mar2023[[6]]),big.mark=",")` cases each month. The end of 2021, was the time when the country was hit by a severe drought that had aggravated even more the already existing vulnerabilities. As a result, several livelihoods were on the brink of Famine [@FRC]. The increased trend reflects the intensified program response at the time, to fit the also increased needs. The results shows, as well, that even after the drought has passed, the magnitude of the admissions remained excessively high, compared to years before the drought, although it shows a downward trend. This suggests that the program response remained high even after. The shallow decline observed in the slope suggests that the magnitude could remain high still in the following years before catching up with the pattern before the drought. This means that in case the program becomes underfunded in the short-term, the trend line might portray a false improvement in the situation (by lowering the magnitude), as this would be the result of many children left untreated due to the scale-down of program capacity.    

As for the **seasonality**, all in all, May-June is the month when the higher peak is recorded, with the increase beginning in May. These patterns tend to be consistent with those identified by @ENN using 8-year SMART surveys conducted by the Somalia Food Security and Nutrition Analysis Unit (FSNAU) in rural areas, however, with some nuances. @ENN identified two peaks in the prevalence of acute malnutrition: May-June and November. One peculiar difference is that in this study, the high peak was recorded in November where the average monthly wasting was estimated above 17.5%; May-June was the second high peak. This current analysis contrasts considerably on this in the opposite way: May-June is the time when the highest peak is observed. This pattern has been consistently noticeable over the years.  

With regard to the second peak, whilst @ENN identified a second peak in November, this analysis' second high peak is January, however with an increase beginning in November. 
On to the trough, this analysis' lower trough is in April every year. @ENN identified a lower trough in August and a second in December. From this analysis findings, August used to be the second lower trough before 2022 with the data aggregated at national-level. However, as of 2022, this pattern has shifted to September, although the extent of the shift is minimal.  

Seasonal patterns exhibit notable variations across livelihood systems. In the *pastoral and agropastoral* systems, the primary peak is consistently observed in May-June, with a secondary peak in January. This secondary peak begins in November for the pastoral system, and in December for the agropastoral system. The lowest trough occurs in April in both systems. These findings align with @ENN, except for a slight deviation in the agropastoral system, where the second peak shifts from November (as identified by @ENN) to December in this analysis.  

In contrast, the patterns in the riverine system differ significantly from those in other livelihood systems. **It is the only system where the primary peak does not occur in May-June**. These patterns also diverge from @ENN observations, which identified the higher peak in November. In this analysis, November consistently represents the lowest trough. The highest peak occurs in December-January, followed by several smaller peaks throughout the year, indicating a more fragmented seasonal pattern.  

::: {.callout-note}
An insteresting finding is that the riverine and pastoral, are they one livelihood systems where there is a peak in March before the lower trough in April.
**Why is that??** *Question to the Nutrition Cluster members.*
:::  

Regarding the *Urban/IDPs livelihood system*, this system exhibits a unique and intriguing pattern compared to the others. While the highest peak consistently occurs in May-June, similar to most other systems (except the riverine), the admissions in January notably align with the trend line (baseline) every year. From February onward, the admissions begin a continuous decline, reaching the common lowest trough in April. This distinct pattern could not be compared to previous studies, as they did not include a specific livelihood category for urban areas and/or IDPs.  
**Why is that??** *Question to the Nutrition Cluster members.*   

::: {.callout-important}
The discrepancies in the months in @ENN 's findings may be related to the fact that the survey data used were collected over a specific time period, while the admissions are reported monthly over the years. This also explains the absence of a trough in April in the author's study, as all surveys were undertook starting in May and then November to December.
:::

### Do these seasonal patterns agree with the seasonality of food security?  
The unveiled AMN seasonal patterns do not agree with or adhere to to food security Jilaal, Gu and Hagaa's season, as it would, theoretically, be expected, but they do agree (to some extent) with the Deyr season.  

  - **Jilaal** is Somalia's extended dry season, with low milk availability and increased livestock migration in pastoralist, in particular. Despite that, during its length, the admissions follow a decreasing behaviour from February with its lower trough in April. This is true for all livelihood system, though there some shallow peaks in March in the Riverine and Pastoral. This raises questions on what non-food security factors could be playing a role.  

  - **Gu**, on the other hand, is Somalia's heavier rainfall, accounting for 70% of production. This is the period where in all livelihood system (except riverine) AMN has its higher peak of admissions. 

  > _The pastoral consumption year starts with the onset of the Gu rainy season (April) when species give birth, and households have access to their own milk a meat and income for other essentials from saleable animals_ - @ENN.

  Even with that, the admissions rise. This seem to indicate the existence of other key drivers in the play that tends to deteriorate, and likely inhibit food utilization. 

## Limitations  
As noted, MAM data was excluded from the analysis due to its variance being influenced by program stockouts, resulting in the study not encompassing GAM data. This raises questions about whether similar patterns would emerge if GAM had been analyzed. Additionally, the SAM reporting system does not differentiate admissions originating from IDPs versus those from other livelihood systems. This poses a significant limitation in understanding the specific patterns among IDPs. Furthermore, the results reveal that SAM admissions follow a distinct seasonality compared to acute food insecurity, with declines in admissions observed during the lean season. Understanding the factors potentially driving this decline would be critical, but this analysis could not explore these aspects further due to data limitations.
